\MakeUppercase{Graph Product Structure for Non-Minor-Closed Classes}

### 2022-03-08 k-planar.tex

Let $G$ be an edge-maximal $1$-plane multigraph with no two parallel edges on the boundary of a single face.  Here, edge-maximal should be taken to mean that, if any two vertices $v$ and $w$ appear on a common face\footnote{The \defin{faces} of an embedded graph $G$ are the connected components of $\R^2\setminus \bigcup_{vw\in E(G)} vw$.  We say that a vertex $v\in V(G)$ appears on a face $F$ if $v$ is contained in the closure of $F$.} $F$, then there is an edge $vw\in E(G)$ that is contained in the boundary of $F$.  (Refer to \cref{one_planar_example}(a,b).) We assume that no two edges incident to a common vertex of $G$ cross each other since, in a 1-plane graph, such a crossing can always be removed by a local modification to obtain an isomorphic 1-plane graph in which the two edges do not cross.\footnote{While this is true for 1-plane graphs it is not true for $k$-plane graphs with $k\ge 3$; the uncrossing operation can increase the number of crossings on a particular edge from $k$ to $2(k-1)$.}

\begin{figure}
  \begin{center}
    \begin{tabular}{c@{\hspace{1cm}}c}
%      \includegraphics{figs/one_planar_example-1} &
%      \includegraphics{figs/one_planar_example-2} \\
%      (a) & (b) \\[1em]
      \includegraphics{figs/one_planar_example-3} &
      \includegraphics{figs/one_planar_example-4} \\
      (a) & (b) 
   \end{tabular}
  \end{center}
%  \caption{Examples of (a)~a $1$-plane graph that is not edge maximal; (b)~an edge-maximal $1$-plane graph $G$; (c)~the kites in $G$; (d)~a plane multigraph $G'$ obtained by removing one spar from each kite of $G$.} 
  \caption{(a)~Kites in an edge-maximal $1$-plane multigraph $G$; (b)~Plane multigraph $G'$ obtained by removing one spar from each kite of $G$.} 
  \label{one_planar_example} 
\end{figure}

To understand the structure of $G$, it is helpful to consider the plane graph $G_0$ obtained by replacing each pair of edges $vw$ and $xy$ involved in a crossing $(p,vw,xy)$ with four edges $vp$, $wp$, $xp$, and $yp$ meeting at a newly added \defin{dummy vertex} $p$. Let $F$ be a face of $G_0$ and let $W:=w_0,\ldots,w_r$ be the facial walk around $F$. By edge maximality, if $W$ contains only non-dummy vertices, then it contains exactly three vertices and $F$ is bounded by three edges of $G$.

If $W$ contains a dummy vertex $w_i$ then neither $w_{i-1}$ nor $w_{i+1}$ is a dummy vertex since if $w_{i-1}$ (respectively $w_{i+1}$) were a dummy vertex then the edge of $G$ that contains $w_{i-1}w_i$ (respectively $w_iw_{i+1}$) is involved in at least two crossings.  Therefore each dummy vertex $w_i$ of $W$ is surrounded by two vertices $w_{i-1},w_{i+1}\in V(G)$.  Since $w_i$ has degree $4$ in $G_0$, $w_{i-1}\neq w_i$.  Since $G$ is edge-maximal the edge $w_{i-1}w_i$ is an edge of $G$ on the boundary of $F$.  Therefore, if $F$ has a dummy vertex $w_i$ in its facial walk $W$, then $W=w_{i-1},w_i,w_{i+1}$ and $F$ is bounded by three edges of $G_0$, each of which is contained in a different edge of $G$.

Therefore, each face of $G$ is one of two types:
\begin{inparaenum}[(i)] 
  \item bounded by three edges of $G$ (none of which are crossed) or 
  \item bounded by one edge of $G$ that is not crossed plus portions of two edges of $G$ that cross each other.
\end{inparaenum}
Consider some crossing $(p,vw,xy)$ in $G$.  The point $p$ at which this crossing occurs is a vertex of $G_0$ that has degree $4$.  There are four faces $F_1,\ldots,F_4$ of $G_0$ that contain $p$.  Each of these four faces of is of Type~(ii), so the boundary of $F_1\cup\cdots\cup F_4$ includes a $4$-cycle $C:=vxwy$ in $G$ whose edges are not involved in any crossings.  We call the subgraph $K:=C\cup\{vw,xy\}$ a \defin{kite} of $G$; see \cref{one_planar_example}(a).  The edges $vw$ and $xy$ are called \defin{spars} of $K$.  The cycle $vxwy$ is called the \defin{sail} of $K$.  Thus any edge that is a spar of a kite $K$ is not part of a sail of any kite $K'$.

@@

By removing one spar from each kite of $G$ we obtain a plane multigraph $G'$ each of whose faces is bounded by three edges; see \cref{one_planar_example}(b).  Observe that, for any spar $xy\in E(G)\setminus E(G')$ that crosses $vw\in E(G')$, $G'$ contains the path $vxw$ (and $vyw$).  It follows that $\dist_{G'}(v,w)\le 2$.

@@

Prior to the current work, the strongest upper bound on the non-repetitive chromatic number of $n$-vertex  $k$-planar graphs was $O(k\log n)$ \cite{dujmovic.morin.ea:layered}.

@@

In this section, we show that $k$-nearest neighbour graphs of point sets in the plane are $O(k^2)$-planar.  For two points $x,y\in\R^2$, let $d_2(x,y)$ denote the Euclidean distance between $x$ and $y$. The $k$-nearest-neighbour graph of a point set $P\subset\R^2$ is the geometric graph $G$ with vertex set $V(G)=P$, where the edge set is defined as follows. For each point $v\in P$, let $N_k(v)$ be the set of $k$ points in $P$ closest to $v$. Then $vw\in E(G)$ if and only if $w\in N_k(v)$ or $v\in N_k(w)$. (The edges of $G$ are straight-line segments joining their endpoints.) See \citep{ProximityGraphs} for a survey of results on $k$-nearest neighbour graphs and other related proximity graphs.

>>>>

In the following, we will assume that no two edges incident to a common vertex of the $1$-plane graph $G$ cross each other since, in a 1-plane graph, such a crossing can always be removed by a local modification to obtain an isomorphic 1-plane graph in which the two edges do not cross.\footnote{While this is true for 1-plane graphs it is not true for $k$-plane graphs with $k\ge 3$; the uncrossing operation can increase the number of crossings on a particular edge from $k$ to $2(k-1)$.}  The \defin{faces} of $G$ are the connected components of $\R^2\setminus \bigcup_{vw\in E(G)} vw$, where we treat each edge $vw$ of $G$ as the image of a simple curve $f_{vw}:[0,1]\to\R^2$.  We say that a vertex $v\in V(G)$ appears on a face $F$ if $v$ is contained in the closure of $F$.

Let $G$ be a $1$-plane multigraph with no two parallel edges on the boundary of a single face.  By Euler's Formula, $G$ has $O(|V(G)|)$ vertices.  We further assume that $G$ is edge-maximal, in the following sense:  if two vertices $v$ and $w$ of $G$ appear on a common face $F$, then there is an edge $vw\in E(G)$ that is contained in the boundary of $F$.  See  \cref{one_planar_example}(b) for an example.  Any $1$-plane multigraph $G$ with at least three vertices can be made edge-maximal by the addition of at most $3|V(G)|-6$ edges since each edge added this way has endpoints in $V(G)$ and does not cross any other edge of $G$.

\begin{figure}
  \begin{center}
    \begin{tabular}{c@{\hspace{1cm}}c}
        \includegraphics{figs/one_planar_example-3} &
        \includegraphics{figs/one_planar_example-4} \\
        (a) & (b)
     \end{tabular}
    \end{center}
    \caption{(a)~Kites in an edge-maximal $1$-plane multigraph $G$; (b)~Plane multigraph $G'$ obtained by removing one spar from each kite of $G$.}
    \label{one_planar_example}
\end{figure}

To understand the structure of $G$, it is helpful to consider the plane graph $G_0$ obtained by replacing each pair of edges $vw$ and $xy$ involved in a crossing $(p,vw,xy)$ with four edges $vp$, $wp$, $xp$, and $yp$ meeting at a newly added \emph{dummy vertex} $p$. Let $F$ be a face of $G_0$ and let $W:=w_0,\ldots,w_{r-1}$ be the facial walk around $F$. In the following paragraphs, subscripts on the vertices in $W$ are implicitly taken modulo $r$. By edge maximality, if $W$ contains only non-dummy vertices, then it contains exactly three vertices and $F$ is bounded by three edges of $G$.

If $W$ contains a dummy vertex $w_i$ then neither $w_{i-1}$ nor $w_{i+1}$ is a dummy vertex since if $w_{i-1}$ (respectively $w_{i+1}$) were a dummy vertex then the edge of $G$ that contains $w_{i-1}w_i$ (respectively $w_iw_{i+1}$) would be involved in at least two crossings.  Therefore each dummy vertex $w_i$ of $W$ is surrounded by two vertices $w_{i-1},w_{i+1}\in V(G)$.  Since $w_i$ has degree $4$ in $G_0$, $w_{i-1}\neq w_{i+1}$.  Since $G$ is edge-maximal the edge $w_{i-1}w_i$ is an edge of $G$ on the boundary of $F$.  Therefore, if $F$ has a dummy vertex $w_i$ in its facial walk $W$, then $W=w_{i-1},w_i,w_{i+1}$ and $F$ is bounded by three edges of $G_0$, each of which is contained in a different edge of $G$.

Therefore each face of $G'$ is a $3$-cycle that contains at most one dummy vertex.  This implies that each face of $G$ is one of two types:
\begin{inparaenum}[(i)]
  \item bounded by three edges of $G$, none of which are crossed; or
  \item bounded by one edge of $G$ that is not crossed plus portions of two edges of $G$ that cross each other.
\end{inparaenum}

Consider some crossing $(p,vw,xy)$ in $G$.  The point $p$ at which this crossing occurs is a vertex of $G_0$ that has degree $4$.  There are four faces $F_1,\ldots,F_4$ of $G_0$ that contain $p$.  Each of these four faces of is of Type~(ii), so the boundary of $F_1\cup\cdots\cup F_4$ includes a $4$-cycle $C:=vxwy$ in $G$ whose edges are not involved in any crossings.  We call the subgraph $K:=C\cup\{vw,xy\}$ a \defin{kite} of $G$; see \cref{one_planar_example}(a).  The edges $vw$ and $xy$ are called \defin{spars} of $K$.  The cycle $vxwy$ is called the \defin{sail} of $K$.  Thus any edge that is a spar of a kite $K$ is not part of a sail of any kite $K'$.

@@

By removing one spar from each kite of $G$ we obtain a plane multigraph $G'$ each of whose faces is bounded by three edges; see \cref{one_planar_example}(b).  Observe that, for any spar $xy\in E(G)\setminus E(G')$ that crosses $vw\in E(G')$, $G'$ contains the path $vxw$ (and $vyw$).  Therefore $\dist_{G'}(v,w)\le 2$.

@@

Prior to the current work, the strongest upper bound on the non-repetitive chromatic number of $n$-vertex  $k$-planar graphs was $O(k\log n)$ \cite{dujmovic.morin.ea:layered}.
=======
%      \includegraphics{figs/one_planar_example-1} &
%      \includegraphics{figs/one_planar_example-2} \\
%      (a) & (b) \\[1em]
%  \caption{Examples of (a)~a $1$-plane graph that is not edge maximal; (b)~an edge-maximal $1$-plane graph $G$; (c)~the kites in $G$; (d)~a plane multigraph $G'$ obtained by removing one spar from each kite of $G$.}

To understand the structure of $G$, it is helpful to consider the plane graph $G_0$ obtained by replacing each pair of edges $vw$ and $xy$ involved in a crossing $(p,vw,xy)$ with four edges $vp$, $wp$, $xp$, and $yp$ meeting at a newly added \defin{dummy vertex} $p$. Let $F$ be a face of $G_0$ and let $W:=w_0,\ldots,w_r$ be the facial walk around $F$. By edge maximality, if $W$ contains only non-dummy vertices, then it contains exactly three vertices and $F$ is bounded by three edges of $G$.
>>>>>>> 432123323ba65fbac04503f0de5e50ec4ebd89ce

If $W$ contains a dummy vertex $w_i$ then neither $w_{i-1}$ nor $w_{i+1}$ is a dummy vertex since if $w_{i-1}$ (respectively $w_{i+1}$) were a dummy vertex then the edge of $G$ that contains $w_{i-1}w_i$ (respectively $w_iw_{i+1}$) woul

@@

In this section, we show that $k$-nearest neighbour graphs of point sets in the plane are $O(k^2)$-planar.  For two points $x,y\in\R^2$, let $d_2(x,y)$ denote the Euclidean distance between $x$ and $y$. The $k$-nearest-neighbour graph of a point set $P\subset\R^2$ is the geometric graph $G$ with vertex set $V(G)=P$, where the edge set is defined as follows. For each point $v\in P$, let $N_k(v)$ be the set of $k$ points i
=======
%      \includegraphics{figs/one_planar_example-1} &
%      \includegraphics{figs/one_planar_example-2} \\
%      (a) & (b) \\[1em]
%  \caption{Examples of (a)~a $1$-plane graph that is not edge maximal; (b)~an edge-maximal $1$-plane graph $G$; (c)~the kites in $G$; (d)~a plane multigraph $G'$ obtained by removing one spar from each kite of $G$.}

To understand the structure of $G$, it is helpful to consider the plane graph $G_0$ obtained by replacing each pair of edges $vw$ and $xy$ involved in a crossing $(p,vw,xy)$ with four edges $vp$, $wp$, $xp$, and $yp$ meeting at a newly added \defin{dummy vertex} $p$. Let $F$ be a face of $G_0$ and let $W:=w_0,\ldots,w_r$ be the facial walk around $F$. By edge maximality, if $W$ contains only non-dummy vertices, then it contains exactly three vertices and $F$ is bounded by three edges of $G$.
>>>>>>> 432123323ba65fbac04503f0de5e50ec4ebd89ce

If $W$ contains a dummy vertex $w_i$ then neither $w_{i-1}$ nor $w_{i+1}$ is a dummy vertex since if $w_{i-1}$ (respectively $w_{i+1}$) were a dummy vertex then the edge of $G$ that contains $w_{i-1}w_i$ (respectively $w_iw_{i+1}$) wouln $P$ closest to $v$. Then $vw\in E(G)$ if and only if $w\in N_k(v)$ or $v\in N_k(w)$. (The edges of $G$ are straight-line segments joining their endpoints.) See \citep{ProximityGraphs} for a survey of results on $k$-nearest neighbour graphs and other related proximity graphs.
